---
title: "TCGA BRCA Data Exploration"
output: html_notebook
---

## Setup

```{r}
library(tidyverse)
library(synapser)
library(stringr)
library(janitor)
library(ggthemes)
library(patchwork)
library(ggalluvial)
library(forcats)
library(annotables)
```

## Data collection

### Clinical data

```{r, warning=FALSE, message=FALSE}
metabric_clinical_df <- read_tsv("../data/raw/METABRIC/METABRIC_discovery.info.tsv") %>% 
  rename_at(vars(contains("NOT_IN_OSLOVAL")), funs(str_replace(., "NOT_IN_OSLOVAL_", "")))
```


Here's a glimpse of the information included in `metabric_clinical_df`:

```{r}
dim(metabric_clinical_df)
summary(metabric_clinical_df)
```


### Expression data

**Note:** I'll use the `annotables` package to augment the provided Entrez gene IDs with some additional field to provide more context: 

+ `ensgene` (which I'll rename to `gene_id` to match the PS-ON data): Ensembl gene ID
+ `symbol`: HGNC gene symbol
+ `biotype`: Protein coding, pseudogene, mitochondrial tRNA, etc.
+ `description`: Full gene name/description

```{r include=FALSE}
metabric_disc_expr_file <- "../data/metabric_disc_expr_df.RData"
metabric_val_expr_file <- "../data/metabric_val_expr_df.RData"
if (!fs::file_exists(metabric_disc_expr_file)) {
  # download and load normalized, batch-corrected expression data
  metabric_disc_expr_df <- read_tsv(
    "../data/raw/METABRIC/METABRIC_discovery.Illumina_HT12.matrix.txt",
    progress = FALSE
  )
  
  # update/augment gene identifiers; subset to protein coding genes ; 
  # remove duplicate IDs  gene_df <- tcga_expr_df %>% 
  select(symbol = Name) %>% 
    filter(symbol != "---") %>% 
    left_join(grch37, by = "symbol") %>% 
    filter(biotype == "protein_coding") %>% 
    dplyr::select(gene_id = ensgene, entrez, symbol, biotype, description) %>% 
    filter(!duplicated(entrez) & !duplicated(gene_id))

  # filter genes
  metabric_disc_expr_df <- metabric_disc_expr_df %>% 
    left_join(gene_df, ., by = c("symbol" = "Name")) %>% 
    group_by(gene_id, symbol) %>% 
    summarise_at(vars(starts_with("MB_")), funs(max))
  
  # make sure gene info lines up with expression data
  gene_df <- gene_df %>% 
    dplyr::filter(gene_id %in% metabric_expr_df$gene_id) %>% 
    arrange(gene_id)
    
  save(metabric_disc_expr_df, file = metabric_disc_expr_file)
  # also save TSV
  # write_tsv(brca_expr_norm_df, 
  #           path = str_replace(brca_expr_norm_file, "RData", "tsv"))
  
  # filter validation data as well
  metabric_val_expr_df <- read_tsv(
    "../data/raw/METABRIC/METABRIC_validation.Illumina_HT12.matrix.txt",
    progress = FALSE
  )
  
  metabric_val_expr_df <- metabric_val_expr_df %>% 
    left_join(gene_df, ., by = c("symbol" = "Name")) %>% 
    group_by(gene_id, symbol) %>% 
    summarise_at(vars(starts_with("MB_")), funs(max))
  
  save(metabric_val_expr_df, file = metabric_val_expr_file)

  save(gene_df, file = "../data/metabric_expr_gene_info.RData")
  # also save TSV
  write_tsv(gene_df, path = "../data/tcga_brca_expr_gene_info.tsv")

  } else {
  load(brca_expr_norm_file)
  load("../data/tcga_brca_expr_gene_info.RData")
}
```

---

## Data formatting

### Clinical data

I'll subset the clinical dataframes to match the selection of samples with expression data, then save the results.

```{r}
metabric_disc_clin_df <- names(metabric_disc_expr_df)[-c(1:2)] %>% 
  tibble(metabric_id = .) %>% 
  left_join(metabric_clinical_df, by = c("metabric_id" = "Array")) %>% 
  mutate(survival_5y = (T / 365.25) >= 5,
         censored = last_follow_up_status %in% c("a", "d-o.c."))

metabric_val_clin_df <- names(metabric_val_expr_df)[-c(1:2)] %>% 
  tibble(metabric_id = .) %>% 
  left_join(metabric_clinical_df, by = c("metabric_id" = "Array")) %>% 
  mutate(survival_5y = (T / 365.25) >= 5,
         censored = last_follow_up_status %in% c("a", "d-o.c."))

metabric_val_masked_clin_df <- metabric_val_clin_df %>% 
  select(-T, -OS_Time)

save(metabric_disc_clin_df, file = "../data/metabric_disc_clin_df.RData")
save(metabric_val_clin_df, file = "../data/metabric_val_clin_df.RData")
save(metabric_val_masked_clin_df, file = "../data/metabric_val_masked_clin_df.RData")
```


```{r}
View(metabric_disc_clin_df)
```




